---
title: "Oil Spills"
description: |
  Working with spatial data to create a map of California oil spills
author:
  - name: Rachel Rhodes
    url: {}
date: 2021-02-20
categories:
  - geospatial data
  - marine
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(tmap)
library(gstat)
library(stars)
library(pals)
library(lubridate)
```

# California oil spills {.tabset}
An exploration of oil spill response in California using data from the Office of Spill Prevention and Response (OSPR) Incident Tracking Database. This database is a statewide oil spill tracking information system that tracks a discharge or threatened discharge of petroleum or other deleterious material into the waters of the state."

**Citation:** Oil Spill Incident Tracking 2008 edition, California Department of Fish and Game, Office of Spill Prevention and Response, July 23, 2009. [https://map.dfg.ca.gov/metadata/ds0394.html](https://map.dfg.ca.gov/metadata/ds0394.html)  

## Overview of Data

#### Data cleanup & prep

The data is collected by OSPR Field Response Team members all over the state and includes soil spills in different environments including marine, fresh water, and on land. The following map explores the total spill incidents recorded for 2008 across the entire state. 


```{r, warning=FALSE, message=FALSE}
## read in CA DFW Oil Spill Incident Tracking
ca_oil <- read_sf("oil_spills",
                  layer = "Oil_Spill_Incident_Tracking_%5Bds394%5D") %>% 
  clean_names()

# st_crs(ca_oil)
# plot(ca_oil)

## Create data frame to quickly see the info stored inside SF 
ca_oil_df <- ca_oil %>% 
  as.data.frame()

## read in CA County Data
ca_counties <- read_sf("ca_counties", 
                       layer = "CA_Counties_TIGER2016") %>% 
  clean_names() %>% 
  select(name, aland)

## check crs to make sure it matches
# st_crs(ca_counties)

## create df to check
ca_counties_df <- ca_counties %>% 
  as.data.frame()

# plot(ca_counties)

## set tmap mode
tmap_mode("view")
```

#### Data Visualization
```{r}
## create interactive map
tm_shape(ca_oil) +
  tm_dots("specificlo", 
          size = 0.04,
          palette=c("cadetblue2", "chartreuse3", "cornflowerblue"))
```



**Figure 1.** Oil spill incidents in California in 2008 by type of environment, includes fresh water, marine, and land.


## Analysis: Inland Spills in 2008

#### Data cleanup & prep

The data includes which county each spill occurs in. The following explores the number of inland spills by county to see if there are any counties that have an especially high number of incidents. Figure 1 highlights that Los Angeles county had the highest number of incidents (340) in 2008 followed by San Mateo county (173).


```{r}
## Filter data to only include inland spills
ca_oil_2008 <- ca_oil %>% 
  mutate(date = ymd(dateofinci)) %>% 
  mutate(year = year(date)) %>% 
  filter(inlandmari == "Inland") %>% 
  count(localecoun) %>% 
  rename(name = localecoun) %>% 
  rename(num_spills = n)

## create data frame to easily look at the data and verify
ca_oil_2008_df <- ca_oil_2008 %>% 
  as.data.frame()

## lets join the two layers together
oil_2008_counties <- ca_counties %>% 
  st_join(ca_oil_2008)

oil_2008_counties_df <- oil_2008_counties %>% 
  as.data.frame()
```

#### Data Visualization
```{r}
ggplot(data = oil_2008_counties)+
  geom_sf(aes(fill = num_spills), color = "white", size = 0.1) +
  scale_fill_gradientn(colors= c("azure3", "cadetblue4","darkcyan", "cyan3"), na.value = "lavenderblush2") +
  theme_minimal() +
  labs (fill = "Number of oil spill incidents")
```


**Figure 2.** The number of oil spill incidents by county in California in 2008 with the highest number of incidents shown in bright blue occurring in Los Angeles county. 